--
title: "Principal Component and Logistic Regression Analysis"
author: "Oscar Saiz Gutierrez"
date: "2024-06-18"
output:
  prettydoc::html_pretty:
    theme: architect
---

```{r message=FALSE, warning=FALSE, results='hide', fig.keep='all'}
library(readr)
library(factoextra)
library(pscl)
library(caret)
library(reshape2)
library(dplyr)
library(FactoMineR)
library(car)
library(nortest)
library(gtsummary)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Data Preparation and Normality Testing

We load the dataset, which includes dietary intake, clinical variables, and disease prevalence for 4900 individuals.

```{r}
data <- read_csv("data/alimentos_nutrientes_4900.csv")
```

To verify statistical assumptions, we test normality for all food (alimento1–131) and nutrient (nutriente1–19) variables using the Anderson–Darling test.

```{r}
for(i in 1:131) {
  alimentos <- ad.test(data[[paste0("alimento", i)]])
}
for(i in 1:19) {
  nutrientes <- ad.test(data[[paste0("nutriente", i)]])
}

TablaNormalidad <- rbind(
  data.frame(
    Variable = paste0("alimento", 1:131),
    Test = "Anderson-Darling",
    p_value = alimentos$p.value,
    Interpretation = ifelse(alimentos$p.value < 0.05, "Not Normal", "Normal")
  ),
  data.frame(
    Variable = paste0("nutriente", 1:19),
    Test = "Anderson-Darling",
    p_value = nutrientes$p.value,
    Interpretation = ifelse(nutrientes$p.value < 0.05, "Not Normal", "Normal")
  )
)
write.csv(TablaNormalidad, "output/normality_results.csv")
```

# 2. Principal Component Analysis (PCA)

We isolate and standardize food and nutrient variables before performing PCA.

```{r}
food_vars <- paste0("alimento", 1:131)
nutrient_vars <- paste0("nutriente", 1:19)
datapca <- data[, c(food_vars, nutrient_vars)]
datapca_scaled <- scale(datapca)
```

```{r}
pca <- prcomp(datapca_scaled)
PVE <- 100 * pca$sdev^2 / sum(pca$sdev^2)
write.csv(data.frame(PC = 1:length(PVE), R2 = PVE), "output/pca_variance.csv")
```

```{r}
fviz_eig(pca, ylim = c(0, 15), addlabels = TRUE, title = "PCA: Explained Variance")
```

We choose PC1 and PC2, which together explain ~14.74% of the total variance.

# 3. PCA Loadings and Contributions

We extract the loadings (rotations) of PC1 and PC2.

```{r}
loadings <- pca$rotation[, 1:2]
write.csv(loadings, "output/pca_loadings.csv")

fviz_contrib(pca, choice = "var", axes = 1, top = 10, addlabels = TRUE)
fviz_contrib(pca, choice = "var", axes = 2, top = 10, addlabels = TRUE)
```

# 4. Descriptive Analysis by PCA Tertiles

We categorize PC1 and PC2 into tertiles to explore associations.

```{r}
data$Tertile_PC1 <- cut(pca$x[, 1], breaks = quantile(pca$x[, 1], probs = seq(0, 1, by = 1/3)), labels = FALSE)
data$Tertile_PC2 <- cut(pca$x[, 2], breaks = quantile(pca$x[, 2], probs = seq(0, 1, by = 1/3)), labels = FALSE)
data$scores_PC1 <- pca$x[, 1]
data$scores_PC2 <- pca$x[, 2]
```

We build summary tables with clinical and sociodemographic variables by tertile.

```{r}
categorical_vars <- c("sexo", "estado_civil", "tabaco", "colesterol", "hdl", "HTA",
                      "hipercolesterolemia", "hipertrigliceridemia", "ECV_prev", "diab_prev")

for (var in categorical_vars) {
  data[[var]] <- as.factor(data[[var]])
}

table_PC1 <- data %>%
  tbl_summary(by = Tertile_PC1,
              statistic = list(all_continuous() ~ "{mean} ± ({sd})")) %>%
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 3))

table_PC2 <- data %>%
  tbl_summary(by = Tertile_PC2,
              statistic = list(all_continuous() ~ "{mean} ± ({sd})")) %>%
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 3))

tbl_merge(list(table_PC1, table_PC2), tab_spanner = c("PC1", "PC2"))
```

# 5. Logistic Regression Models

We predict diabetes prevalence (`diab_prev`) using PCA tertiles and key covariates.

```{r}
data$diab_prev <- factor(data$diab_prev)
data$sexo <- factor(data$sexo)

model1 <- glm(diab_prev ~ Tertile_PC1 + Tertile_PC2 + sexo + edad + hdl,
              data = data, family = "binomial")

model2 <- glm(diab_prev ~ Tertile_PC1 + Tertile_PC2 + sexo + edad + altura + peso +
                tabaco + colesterol + hdl + HTA + hipercolesterolemia +
                hipertrigliceridemia + ECV_prev,
              data = data, family = "binomial")
```

```{r}
summary(model1)
summary(model2)

exp(coef(model1))
exp(coef(model2))

1 - (model1$deviance / model1$null.deviance)
1 - (model2$deviance / model2$null.deviance)
```

```{r}
tbl_merge(list(
  tbl_regression(model1, exponentiate = TRUE),
  tbl_regression(model2, exponentiate = TRUE)
), tab_spanner = c("Model 1", "Model 2"))
```
